Conflict Driven Evolution and Urban Development: \\ Evidence from the Holy Roman Empire\footnote{ I thank Lara Andermann, Leonie Bielefeld, Daniel Bursian, Davide Cantoni, Emilio Esguerra, Filip Milojevic, Johannes Vorbach and Matthias Weigand for helpful comments and feedback. }

### 2023-06-26 paper/paper.tex
#29 discuss attenuation bias

\documentclass[11pt, a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[left=3cm,right=3cm,top=2.5cm,bottom=2.5cm]{geometry}  
% set 11pt and right=6cm for submission

@@

\newcommand{\decoline}{% pretty "scene change" macro
  \par
  \vspace{\baselineskip}
  \hfill
  {\Large*}
  \hfill
  \vspace{\baselineskip}
\par}

@@

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win with state power, and state power is simply free resources, i.e. resources above subsistence. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote
{
The basic idea of Darwinian selection has previously been borrowed by economists - for example by \cite{galor2002} who argue that humans are subject to selection and accumulate (economically) favourable traits over the course of history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that at some point the European rich began to out-breed the poor and passed their favourable traits on to their offspring.
}

@@

The closest relative to this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Further, they show that this long-run benefit is driven by improved state quality after switching. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict. They use a dataset with a bigger geographical scope (all of Europe) but lower resolution: they use urban population from \cite{bairoch1988} as a proxy for development, which is available for fewer cities and fewer years.

@@

To sum these results up, I have found fairly robust non-evidence for the hypothesis that selection for state power drove city growth in the Holy Roman Empire between 1300 and 1800.

>>>>

\documentclass[11pt, a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[left=2cm,right=6cm,top=2.5cm,bottom=2.5cm]{geometry}  
% set 11pt and right=6cm for submission

@@

\newcommand{\decoline}{% pretty "scene change" macro
  \par
  \vspace{0.33\baselineskip}
  \hfill
  {\large* * *}
  \hfill
  \vspace{0.33\baselineskip}
\par}

@@

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win them with state power, and state power is simply aggregate production in excess of the subsistence level. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote
{
The basic idea of Darwinian selection has previously been borrowed by economists - for example by \cite{galor2002} who argue that humans are subject to selection and accumulate (economically) favourable traits over the course of history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that at some point the European rich began to out-breed the poor and passed their favourable traits on to their offspring.
}

@@

The closest relative of this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Further, they show that this long-run benefit is driven by improved state quality after switching. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict. They use a dataset with a bigger geographical scope (all of Europe) but lower resolution: they use urban population from \cite{bairoch1988} as a proxy for development, which is available for fewer cities and fewer years.

@@

To summarize these results: I find robust non-evidence for the hypothesis that selection for state power drove city growth in the Holy Roman Empire between 1300 and 1800. The question is, can we trust this non-finding?

The key economic assumption underlying the identification strategy for the effect of state power was that states that acquire a city by conquest are more powerful on average than states that acquire a city by succession. If this assumption does not hold, then the difference between the treatment effect of a conquest and the treatment effect of a succession is just that: a heterogeneity with no obvious meaning.

I justify this assumption with the fact that a conqueror is, by definition, the winner of an armed conflict. State power is not the only determinant of winning armed conflicts, though: luck plays a central role, as do many other time-varying factors for which I cannot control. If the correlation between state power differentials (the independent variable whose effects we want to measure) and winning conflicts (how we measure it) is small then we essentially have measurement error in the independent variable. This leads to attenuation bias. The estimate for $\theta_conquest$ has the expected positive sign in my baseline regression --- it just lacks statistical significance. If we could regress differences in construction trends directly on observed differences in state power, we would sidestep this measurement error and likely find larger and significant effects.

On the flip side, winning armed conflicts is also not the only consequence of state power. 
That the 
It could even be the opposite: while successions are about familial ties on a superficial level, how this network of familial ties links different lineages is not necessarily uncorrelated with state power. It is a possibility that

On top of this basic assumption about the interpretation of the independent variable come econometric assumptions.

External validity
